Shifting paradigms in Ocean Color: Bayesian Inference for Uncertainty-Aware Chlorophyll Estimation

### 2025-04-10 manuscript/manuscript.qmd
finished introduction draft

#### Historical Context of Chlorophyll Algorithms
Satellite ocean color observations have long been fundamental for monitoring marine ecosystems, as they enable the global estimation of chlorophyll‑a—a key indicator of phytoplankton biomass and ocean productivity. Early empirical algorithms, notably developed by O’Reilly et al. [@oreilly1998; @oreilly2000], established the $OCx$ family (where $x$ denotes the number of bands used) of polynomial regression models. These models relate blue-to-green reflectance ratios (after log‑transformation) to in situ chlorophyll‑a, employing either straight band ratios (BR) or maximum band ratios (MBR)—the latter selecting the highest available blue-to-green ratio for any given observation as input to a high‑order polynomial. These formulations have served as the operational foundation for chlorophyll‑a products across a broad range of satellite ocean color sensors—from the pioneering Coastal Zone Color Scanner (CZCS) through SeaWiFS, MODIS, and MERIS to more recent missions—offering a straightforward and robust approach for Case‑1 waters. However, their performance is more limited in optically complex Case‑2 waters and remains sensitive to atmospheric correction errors.

Subsequent refinements were introduced to address these deficiencies. For example, Hu et al. [@hu2012novel] proposed a Color Index (CI) formulation that employs a band‑difference approach to reduce sensitivity to residual atmospheric errors and instrument noise, with further improvements enhancing inter‑sensor agreement [@hu2019improving]. The increasing availability of calibration data (e.g., [@Valente2015]) and ongoing algorithmic improvements have led to the development of additional variants of the $OCx$ algorithms—specifically, the OC5 and OC6 formulations. O’Reilly and Werdell [@oreilly2019] maintain that OC5 extends the spectral basis by incorporating the 412 nm band, thereby exploiting its strong signal in clear, oligotrophic waters, while OC6 replaces the traditional denominator with the mean of the 555 and 670 nm reflectances, improving the dynamic range at low chlorophyll concentrations. In total, [@oreilly2019] propose 65 versions of BR/MBR $OCx$ type algorithms for 25 sensors—on average, two or more variants per sensor. With this arsenal, it is hoped, researchers are better equipped to address the wide array of bio‑optical environments encountered in global ocean color applications.

#### Limitations of Existing Approaches
Regrettably, the development of traditional ocean color algorithms relies on a fundamental statistical error. The error is the mistaken assumption that the relative frequencies observed in a finite sample can be taken as precise estimates of the underlying true probabilities. In other words, it is the error of conflating  empirical frequencies — derived from a limited, noisy set of observations — with the long-run probability that would be obtained under an infinite number of trials, thereby ignoring the intrinsic uncertainty and variability that come with finite sampling. The fallout has and continues to be extremely detrimental. Traditional, frequentist approaches, where the variability in sample frequencies is erroneously treated as negligible, yield overconfident inferences, and allegedgly significant findings that do not hold water, in time leading to replication crises.

In statistical modeling, the goal is to assess how well a hypothesized model $M$ represents the relationship between independent and dependent variables based on collected data $D$. Frequentist methods focus exclusively on the likelihood $p(D|M)$, estimating model parameters by maximizing this probability. This approach assumes that, with a sufficiently large dataset, the chosen parameters reliably reflect the true system. However, in practice, any finite dataset carries inherent uncertainty and noise. As a consequence, models developed under this framework tend to be overconfident---yielding point estimates that do not adequately capture the variability in in situ observations---and may not generalize well to new data. Ad hoc attempts to generate uncertainty, such as bootstrapping, further compound the problem by resampling the flawed, finite observations and thus amplifying any existing biases. This overconfidence, coupled with unverified assumptions presented as objectivity, can lead to misleading inferences and contribute to replication challenges. The historical process by which these assumptions have been enshrined is fascinating in its own right, and we invite interested readers to consult works such as McGrayne [@mcgrayne] and Clayton [@clayton].

As elaborated extensively by Jaynes [@jaynes2003probability], and more recently by practitioners like Gelman [@gelman2013bayesian], and McElreath [@mcelreath2015statistical], probability is best understood as a degree of knowledge rather than as a long-run frequency. Accordingly in this Bayesian framework, the prior $p(M)$ encodes our current state of knowledge about the model's parameters before $D$ is observed. This knowledge can be elicited from experts, or if none exixts weakly informative priors can be used. Note that whether $M$ is based on empirically observed relationships, analytical understanding, or something in between is irrelevant to its integration into this framework. Once the data $D$ are collected, Bayes' theorem combines the likelihood $p(D|M)$ with the prior $p(M)$ to yield  the posterior distribution $p(M|D)$, which represents an updated level of knowledge about the model's parameters. Crucially this posterior can then serve as the new prior when new data become available, allowing for sequential integration of new information. The by-product of this approach is uncertainty estimates that more accurately reflect the complexities of the underlying system as currently understood. In effect, rather than relying on single point estimates based on $p(D|M)$, the posterior $p(M|D)$ can be mined to retrieve a range of plausible parameter values through credible intervals offering a more transparent and robust basis for inference. This principled integration of prior knowledge with observed data is central to our approach, as detailed further in the Supplementary Material.

In the Bayesian framework, the prior 
p(M) encodes our current state of knowledge about the model's parameters before any data are observed. This knowledge can be elicited from experts or determined experimentally when expert opinion is unavailable. Once data 
D are collected, Bayes’ theorem combines the likelihood 
p(D∣M) with the prior to yield the posterior distribution 
p(M∣D), which represents our updated belief about the parameters. Crucially, this posterior can then serve as the new prior when new data become available, allowing for seamless sequential updating. In this way, whether the model is derived empirically, analytically, or somewhere in between, it is treated as a hypothesis whose parameters are updated in light of new data. This principled integration of prior knowledge with observed data is central to our approach, as detailed further in the Supplementary Material.

#### Motivation for Bayesian Tree-Based Models
To address these limitations, we explore Bayesian tree-based modeling – specifically, Bayesian Additive Regression Trees (BART) – as a novel approach for ocean color chlorophyll retrieval. BART offers several advantages for this application:

Direct Bayesian inference: As a Bayesian ensemble method, BART directly targets the posterior distribution $p(Chl | Rrs)$ rather than a purely empirical fit. The model inherently provides a probabilistic mapping from reflectance to chlorophyll. Rather than producing a single best-fit value, BART returns a posterior mean and credible interval for each prediction, naturally quantifying predictive uncertainty.

Uncertainty quantification: The BART framework includes uncertainty in its predictions by design. Every estimated chlorophyll value comes with a confidence range (e.g., 95% credible interval) reflecting model uncertainty and observational noise. This is a key improvement over deterministic algorithms, enabling more reliable use of the data for scientific and operational purposes where understanding uncertainty is crucial e.g. for forecasting trends.

Interpretability: Being a sum-of-trees model, BART can be interrogated to understand the influence of each input variable. Tools such as partial dependence plots allow us to visualize the relationship between each spectral band and the predicted chlorophyll, marginalizing over other bands. In contrast to high-order polynomial coefficients, these tree-based partial dependencies are intuitively interpretable. For instance, one can observe how changes in a particular waveband’s reflectance (holding others average) drive the chlorophyll estimate, revealing any thresholds or non-linear responses.

BART’s modeling approach is non-parametric and highly flexible, which means it can capture complex, non-linear relationships between multi-spectral Rrs inputs and chlorophyll without imposing a rigid functional form. At the same time, its Bayesian regularization – which shrinks the contribution of each individual tree – helps prevent overfitting even with a rich model structure. This balance between flexibility and regularization is well-suited to global ocean color data, where the true Rrs–Chl relationship is known to be non-linear and context-dependent, yet a model must generalize across diverse water types.

The interpretability of the BART model is illustrated by partial dependence analysis on the six input Rrs bands (centered approximately at 412, 443, 490, 510, 555, and 670 nm). The partial dependence plots reveal nuanced spectral–chlorophyll relationships and clear transitions in the dominant wavelength influence across different chlorophyll regimes. For example, the shorter blue wavelengths (412–443 nm) show a strong inverse influence on predicted $log(Chl_a)$ (higher blue reflectance corresponds to lower chlorophyll) that gradually saturates in the most oligotrophic waters, consistent with the leveling off of blue-to-green ratios at very low Chl. The green band (~555 nm) exhibits a more complex, non-monotonic effect: at low reflectance (clear water) its influence on the Chl estimate is minimal, but it becomes increasingly important through intermediate reflectance ranges – reflecting the band-difference signal exploited by the CI algorithm – and then diminishes for very high chlorophyll where green reflectance tends to flatten. In contrast, the red band (~670 nm) has virtually no effect on the model’s predictions until a threshold is reached at elevated chlorophyll concentrations, after which its influence steeply increases. This behavior aligns with optical expectations, since red reflectance is negligible in low-Chl waters but rises sharply once phytoplankton absorption in the blue saturates and biomass is high. These partial dependence results indicate that the BART model automatically learns the piecewise spectral logic that oceanographers often handle via separate algorithms (blue/green ratios for low-to-moderate Chl, red bands for high Chl). The ability to capture such regime-dependent responses in a single unified model, and to visualize them, underscores the interpretability and scientific insight provided by the BART approach.

Statement of Contribution
In this study, we develop and demonstrate a new global chlorophyll retrieval model based on Bayesian Additive Regression Trees implemented in PyMC. We train the BART model on a large, standardized dataset of satellite remote-sensing reflectance (Rrs) spectra matched with in situ chlorophyll measurements, using log₁₀-transformed Chl-a as the response to stabilize variance. The resulting model is applied globally to produce chlorophyll-a estimates from multi-spectral satellite data, with associated uncertainty estimates for each prediction. We show that this Bayesian tree-based model can serve as a general-purpose ocean color algorithm that is sensor-agnostic (provided reflectances are harmonized to common wavebands), interpretable, and uncertainty-aware. Unlike conventional empirical algorithms, the BART approach allows users to examine the inferred Rrs–Chl relationships and trust the model’s performance across regimes, while also quantifying confidence in each retrieval. This work thus contributes a novel methodological advance to satellite ocean color science: a unified chlorophyll retrieval model that marries the strengths of empirical algorithms (global applicability and simplicity) with the benefits of modern Bayesian machine learning (flexibility, interpretability, and rigorous uncertainty quantification). Our introduction of BART for global chlorophyll prediction opens the door for more robust monitoring of ocean biogeochemistry and improved integration of ocean color data into scientific and management applications.

>>>>

#### Historical Context of Chlorophyll Algorithms
Satellite ocean color observations have long been fundamental for monitoring marine ecosystems, as they enable the global estimation of chlorophyll‑a ($Chl_a$) —a key indicator of phytoplankton biomass and ocean productivity. Early empirical algorithms, notably developed by O’Reilly et al. [@oreilly1998; @oreilly2000], established the $OCx$ family (where $x$ denotes the number of bands used) of polynomial regression models. These models relate blue-to-green reflectance ratios (after log‑transformation) to in situ $Chl_a$, employing either straight band ratios (BR) or maximum band ratios (MBR)—the latter selecting the highest available blue-to-green ratio for any given observation as input to a high‑order polynomial. These formulations have served as the operational foundation for chlorophyll‑a products across a broad range of satellite ocean color sensors—from the pioneering Coastal Zone Color Scanner (CZCS) through SeaWiFS, MODIS, and MERIS to more recent missions—offering a straightforward and robust approach for Case‑1 waters. However, their performance is more limited in optically complex Case‑2 waters and remains sensitive to atmospheric correction errors.

Subsequent refinements were introduced to address these deficiencies. For example, Hu et al. [@hu2012novel] proposed a Color Index (CI) formulation that employs a band‑difference approach to reduce sensitivity to residual atmospheric errors and instrument noise, with further improvements enhancing inter‑sensor agreement [@hu2019improving]. The increasing availability of calibration data (e.g., [@Valente2015]) and ongoing algorithmic improvements have led to the development of additional variants of the $OCx$ algorithms—specifically, the OC5 and OC6 formulations. O’Reilly and Werdell [@oreilly2019] maintain that OC5 extends the spectral basis by incorporating the 412 nm band, thereby exploiting its strong signal in clear, oligotrophic waters, while OC6 replaces the traditional denominator with the mean of the 555 and 670 nm reflectances, with the aim of improving the dynamic range at low chlorophyll concentrations. In total, [@oreilly2019] propose 65 versions of BR/MBR $OCx$ type algorithms for 25 sensors—on average, two or more variants per sensor. With this arsenal, it is hoped, researchers are better equipped to address the wide array of bio‑optical environments encountered in global ocean color applications.

#### Limitations of Existing Approaches
Regrettably, the development of traditional ocean color predictive models relies on a fundamental statistical error that plagues most of observational science today; that of conflating sampling probability, with inferential probability. Consider data $D$ that include predictive features (e.g. Remote sensing reflectance - Rrs) and prediction targets (e.g. $Chl_a$ or phytoplankton absorption), and a model $M$ (e.g. OCx) hypothesized to adequately represent the statistical association between the two. The sampling probability $p(D|M)$ also referred to as the likelihood, is the probability of $D$ conditioned on $M$ being "true". The common model fitting practice is to maximize $p(D|M)$ - likelihood maximization - achieved by tuning $M$'s parameters. The unspoken assumption is that making $M$ fit $D$ as well as possible yields the best version of $M$ given $D$ i.e.  maximizing $p(M|D)$, the inferential probability. This sneaky swapping from $p(D|M)$ to $p(M|D)$ is a violation of the rules of conditional probability, Bayes' theorem in particular. Jaynes[@jaynes2003] and later Clayton through his highly readable account [@clayton] are some of the more notable whistleblowers. Occasionally, in simple  or well-behaved problems, with a large amount of data and no prior information available the maxima of both $p(M|D)$ and $p(D|M)$ can indeed converge on the same answer. Yet there is no guarantee predictability for this outcome however, and the fallout of this abuse is non-negligible, most noticeably in terms of models that don't generalize well, ad-hoc uncertainty estimation attempts, and more broadly a replication crisis.

Some researchers have attempted to deal with these issues. [@seegers2018] e.g. have proposed alternative metrics to circumvent the inadequate assumptions of the frequentist approach. Others have tried to go a step futher and switch to the Bayesian paradigm. E.g. [ @frouin2013] have proposed a Bayesian inversion scheme for atmospheric correction. [@shi2015] proposed a probabilistic method to merge data from different sensors. [@craig2019] have proposed a Bayesian neural network approach using Hamiltonian Monte Carlo sampling to retrieve Inherent Optical Properties (IOP) from Top-of-the-atmosphere (TOA) radiance. [@werther2022] used Monte-Carlo dropout as deep neural net training method to obtain prediction uncertainty. [@erickson2023] have proposed using conjugate Gaussian prior and likelihood to predict IOPs using GIOP as a forward model. [@hammout2024] have proposed a Bayesian Neural Network trained using Stochastic Variational Inference to predict $Chl_a$ from ocean color observations. Yet most of these approaches retain variable levels of frequentism by applying only part of what is commonly referred to as the Bayesian workflow[@betancourt2019].

In this paper we propose a modified workflow to Bayesian modeling for Ocean Color emote sensing. We propose that this workflow include; (1) one or more model building, (2) prior formulation via expert knowledge elicitation and/or data-free simulation, (3) fitting model to data, (4) model validation, comparison and selection, (5) interpretation of results via posterior analysis, (6) prediction on new observation, (7) update the posterior when new labeled data becomes available. As illustration we recast some of the pre-existing $OCx$ models into their probabilistic version. To illustrate model comparison and selection we also build and evaluate Bayesian Additive Tree Regregssion  (BART; [@chipman2010bart]) type model.

### Methods

#### Prior Elicitation and Model Formulation

#### Data Preprocessing

####

\section{Data Preprocessing}

Data for this study were acquired from multiple satellite ocean color sensors and corresponding in situ chlorophyll-\(a\) measurements obtained from sources such as the NASA Bio-Optical Marine Algorithm Data set (NOMAD) and the compilation by Valente et al. (2015). To ensure consistency across sensors, the spectral reflectance data (\(R_{rs}\)) were interpolated as needed to common wavelength centers.

For the empirical \(OCx\) formulation, blue-to-green band ratios were computed for each observation. In particular, the maximum band ratio (MBR) was determined by taking the highest value among the available blue-band ratios (e.g., \(Rrs(443)/Rrs(555)\), \(Rrs(490)/Rrs(555)\), and \(Rrs(510)/Rrs(555)\)). This maximum value was then log-transformed:
$$
\log R = \log_{10}\left(\frac{R_{rs}(\lambda_{\text{blue}})}{R_{rs}(555)}\right).
$$

For the Color Index (CI) formulation of Hu et al. (2012), the CI was calculated as:
$$
\text{CI} = R_{rs}(555) - \left[\,R_{rs}(443) + \frac{555-443}{670-443}\Bigl(R_{rs}(670) - R_{rs}(443)\Bigr)\,\right],
$$
and the corresponding in situ chlorophyll-\(a\) concentrations were log-transformed:
$$
\log \text{Chl} = \log_{10}(\text{Chl}).
$$

These transformations standardize the data to a common scale, ensuring that variability is appropriately captured for subsequent regression and uncertainty quantification. Detailed descriptions of the interpolation methods and quality control procedures are provided in the Supplementary Material.

## Statement of Contribution
In this study, we develop and demonstrate a new global chlorophyll retrieval model based on Bayesian Additive Regression Trees implemented in PyMC. We train the BART model on a large, standardized dataset of satellite remote-sensing reflectance (Rrs) spectra matched with in situ chlorophyll measurements, using log₁₀-transformed Chl-a as the response to stabilize variance. The resulting model is applied globally to produce chlorophyll-a estimates from multi-spectral satellite data, with associated uncertainty estimates for each prediction. We show that this Bayesian tree-based model can serve as a general-purpose ocean color algorithm that is sensor-agnostic (provided reflectances are harmonized to common wavebands), interpretable, and uncertainty-aware. Unlike conventional empirical algorithms, the BART approach allows users to examine the inferred Rrs–Chl relationships and trust the model’s performance across regimes, while also quantifying confidence in each retrieval. This work thus contributes a novel methodological advance to satellite ocean color science: a unified chlorophyll retrieval model that marries the strengths of empirical algorithms (global applicability and simplicity) with the benefits of modern Bayesian machine learning (flexibility, interpretability, and rigorous uncertainty quantification). Our introduction of BART for global chlorophyll prediction opens the door for more robust monitoring of ocean biogeochemistry and improved integration of ocean color data into scientific and management applications.

### 2025-04-10 manuscript/manuscript.tex
finished introduction draft

Unfortunately, the development process for traditional ocean color
algorithms is founded on a statistical fallacy. Frequentism---the
classical approach---assumes that observations are random samples drawn
from the studied phenomena. These samples' probabilities are considered
the observed frequencies that capture the the true - albeit noised up
and potentially biased - behavior of the system. Under this framework,
if a phenomenon is hypothesized to be represented by a model \(M\) and
data \(D\) is collected, then the model's representativeness is
evaluated solely through the likelihood \(p(D∣M)\), which measures the
probability of observing the data given that \(M\) is true. The fallacy
arises from the mistaken belief that, with sufficiently large amounts of
data, one can simply ``flip'' the likelihood to obtain \(p(M∣D)\),
thereby yielding a probabilistic statement about \(M\). Compounding this
error is the imposition of numerous unverified assumptions---disguised
as objectivity---that lead frequentists to set the parameters of \(M\)
by maximizing \(p(D∣M)\), under the false premise that this provides the
best possible representation of the model. In essence, this fallacy
leads to a confusion between the sampling probability \(p(D∣M)\) and the
inferential probability \(p(M∣D)\); the latter when correctly obtained
quantifies the plausibility of the hypothesis represented by \(M\) -
precisely what we seek. The historical dimensions of this frequentist
illusion are fascinating in their own right, and we invite interested
readers to consult works such as McGrayne (\textbf{mcgrayne?}) and
Clayton (\textbf{clayton?}).

Bayesianism, Frequentism's archrival, provides the framoework to
establish the plausibility of \(M\) using observations through the
reinterpretation of how probability should be construed. As elaborated
extensively by Jaynes (\textbf{jaynes2003probability?}), and more
recently by practitioners like Gelman (\textbf{gelman2013bayesian?}),
and McElreath (\textbf{mcelreath2015statistical?}), probability is best
understood as a statement of our knowledge rather than as an inherent
property of the data. the Bayesian perspective treats the data, once
observed, as fixed and the model parameters as latent random variables
the uncertainty of which can---and should---be explicitly modeled via
prior distributions. Prior distributions are probabilistic constructs
that help workers encode prior knowledge. Priors can represents
knowledge elicited from subject matter experts, assumptions held by the
modeler or a complete absence thereof.

This conceptual framework dissolves the false dichotomy between
empirical and semi‑analytical approaches. Both are, in essence,
different hypotheses or approximations about how ocean color data are
generated; they merely provide alternative expressions of the likelihood
\(p(Chl | Rrs)\) to be updated in light of new data. By adopting a
Bayesian paradigm, our approach integrates prior knowledge with observed
data to yield a full posterior predictive distribution---offering not
only improved predictive performance but also principled uncertainty
quantification. In doing so, it circumvents the need for the
proliferation of narrowly tuned, sensor-specific variants (65 versions
for 25 sensors, as documented by O'Reilly and Werdell
(\textbf{OREILLY2019?})) that are symptomatic of ad hoc calibration
processes. Detailed methodological comparisons and additional model
refinements are provided in the Supplementary Material.

@@

Statement of Contribution In this study, we develop and demonstrate a
new global chlorophyll retrieval model based on Bayesian Additive
Regression Trees implemented in PyMC. We train the BART model on a
large, standardized dataset of satellite remote-sensing reflectance
(Rrs) spectra matched with in situ chlorophyll measurements, using
log₁₀-transformed Chl-a as the response to stabilize variance. The
resulting model is applied globally to produce chlorophyll-a estimates
from multi-spectral satellite data, with associated uncertainty
estimates for each prediction. We show that this Bayesian tree-based
model can serve as a general-purpose ocean color algorithm that is
sensor-agnostic (provided reflectances are harmonized to common
wavebands), interpretable, and uncertainty-aware. Unlike conventional
empirical algorithms, the BART approach allows users to examine the
inferred Rrs--Chl relationships and trust the model's performance across
regimes, while also quantifying confidence in each retrieval. This work
thus contributes a novel methodological advance to satellite ocean color
science: a unified chlorophyll retrieval model that marries the
strengths of empirical algorithms (global applicability and simplicity)
with the benefits of modern Bayesian machine learning (flexibility,
interpretability, and rigorous uncertainty quantification). Our
introduction of BART for global chlorophyll prediction opens the door
for more robust monitoring of ocean biogeochemistry and improved
integration of ocean color data into scientific and management
applications.

>>>>

Regrettably, the development of traditional ocean color algorithms
relies on a fundamental statistical error. This error originates in the
conflation of empirical frequencies --- derived from a limited, noisy
set of observations --- with the theoretical relative frequencies that
would be observed over an infinite number of trials, thereby ignoring
the intrinsic uncertainty and variability inherent in finite sampling.
The fallout has and continues to be extremely detrimental to
observational science. Traditional, frequentist approaches, where the
variability in sample frequencies is treated as negligible, yield
overconfident inferences, and allegedgly significant findings that do
not hold water, in time leading to replication crises.

In statistical modeling, the goal is to assess how well a hypothesized
model \(M\) represents the relationship between independent and
dependent variables based on collected data \(D\). Frequentist methods
focus exclusively on the likelihood \(p(D|M)\), estimating model
parameters by maximizing this probability. This approach assumes that,
with a sufficiently large dataset, the chosen parameters reliably
reflect the true system. However, in practice, any finite dataset
carries inherent uncertainty and noise. As a consequence, models
developed under this framework tend to be overconfident - yielding point
estimates that do not adequately capture the variability of
\emph{in-situ} observations - and may not generalize well to new data.
Ad hoc attempts to generate uncertainty, such as bootstrapping, further
compound the problem by resampling the flawed, finite observations and
thus amplifying any existing biases. This overconfidence, coupled with
unverified assumptions presented as objectivity, can lead to misleading
inferences and contribute to replication challenges. The historical
process by which these assumptions have been enshrined is fascinating in
its own right, and we invite interested readers to consult works such as
McGrayne (\textbf{mcgrayne?}) and Clayton (\textbf{clayton?}).

As elaborated extensively by Jaynes (\textbf{jaynes2003probability?}),
and more recently by practitioners like Gelman
(\textbf{gelman2013bayesian?}), and McElreath
(\textbf{mcelreath2015statistical?}), probability is best understood as
a degree of knowledge rather than as a long-run frequency. Accordingly
in this Bayesian framework, the prior \(p(M)\) encodes our current state
of knowledge about the model's parameters before \(D\) is observed. This
knowledge can be elicited from experts, or if none exixts weakly
informative priors can be used. Note that whether \(M\) is based on
empirically observed relationships, analytical understanding, or
something in between is irrelevant to its integration into this
framework. Once the data \(D\) are collected, Bayes' theorem combines
the likelihood \(p(D|M)\) with the prior \(p(M)\) to yield the posterior
distribution \(p(M|D)\), which represents an updated level of knowledge
about the model's parameters. Crucially this posterior can then serve as
the new prior when new data become available, allowing for sequential
integration of new information. The by-product of this approach is
uncertainty estimates that more accurately reflect the complexities of
the underlying system as currently understood. In effect, rather than
relying on single point estimates based on \(p(D|M)\), the posterior
\(p(M|D)\) can be mined to retrieve a range of plausible parameter
values through credible intervals offering a more transparent and robust
basis for inference. This principled integration of prior knowledge with
observed data is central to our approach, as detailed further in the
Supplementary Material.

In this paper we recast a number of OCx-like algorithms in the Bayesian
framework. We also explore a different kind of

@@

\section{Data Preprocessing}

Data for this study were acquired from multiple satellite ocean color
sensors and corresponding in situ chlorophyll-(a) measurements obtained
from sources such as the NASA Bio-Optical Marine Algorithm Data set
(NOMAD) and the compilation by Valente et al.~(2015). To ensure
consistency across sensors, the spectral reflectance data ((R\_\{rs\}))
were interpolated as needed to common wavelength centers.

For the empirical (OCx) formulation, blue-to-green band ratios were
computed for each observation. In particular, the maximum band ratio
(MBR) was determined by taking the highest value among the available
blue-band ratios (e.g., (Rrs(443)/Rrs(555)), (Rrs(490)/Rrs(555)), and
(Rrs(510)/Rrs(555))). This maximum value was then log-transformed: \[
\log R = \log_{10}\left(\frac{R_{rs}(\lambda_{\text{blue}})}{R_{rs}(555)}\right).
\]

For the Color Index (CI) formulation of Hu et al.~(2012), the CI was
calculated as: \[
\text{CI} = R_{rs}(555) - \left[\,R_{rs}(443) + \frac{555-443}{670-443}\Bigl(R_{rs}(670) - R_{rs}(443)\Bigr)\,\right],
\] and the corresponding in situ chlorophyll-(a) concentrations were
log-transformed: \[
\log \text{Chl} = \log_{10}(\text{Chl}).
\]

These transformations standardize the data to a common scale, ensuring
that variability is appropriately captured for subsequent regression and
uncertainty quantification. Detailed descriptions of the interpolation
methods and quality control procedures are provided in the Supplementary
Material.

\section{Statement of Contribution}\label{statement-of-contribution}

In this study, we develop and demonstrate a new global chlorophyll
retrieval model based on Bayesian Additive Regression Trees implemented
in PyMC. We train the BART model on a large, standardized dataset of
satellite remote-sensing reflectance (Rrs) spectra matched with in situ
chlorophyll measurements, using log₁₀-transformed Chl-a as the response
to stabilize variance. The resulting model is applied globally to
produce chlorophyll-a estimates from multi-spectral satellite data, with
associated uncertainty estimates for each prediction. We show that this
Bayesian tree-based model can serve as a general-purpose ocean color
algorithm that is sensor-agnostic (provided reflectances are harmonized
to common wavebands), interpretable, and uncertainty-aware. Unlike
conventional empirical algorithms, the BART approach allows users to
examine the inferred Rrs--Chl relationships and trust the model's
performance across regimes, while also quantifying confidence in each
retrieval. This work thus contributes a novel methodological advance to
satellite ocean color science: a unified chlorophyll retrieval model
that marries the strengths of empirical algorithms (global applicability
and simplicity) with the benefits of modern Bayesian machine learning
(flexibility, interpretability, and rigorous uncertainty
quantification). Our introduction of BART for global chlorophyll
prediction opens the door for more robust monitoring of ocean
biogeochemistry and improved integration of ocean color data into
scientific and management applications.
